Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression

### 2018-11-21 paper/paper.tex
Minor edits to most of the paper.

%\title{Surrogate waveform model using Gaussian process regression for an aligned-spin effective one body model of binary neutron star inspirals}
\title{Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression}

@@

%Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters. Unfortunately, current implementations take several minutes to a day to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate of this model that requires less than 0.3~s to evaluate for a waveform with a starting frequency of 10~Hz (0.04~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods. \red{[[Mismatch is ...]] [[It can be used in arbitrary frequency range by matching to analytic TaylorF2.]] [[Improvement to previously published EOB model [cite] by including spin-induced quadrupole moment and prescription for waveform tapering at merger.]] }

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as the recently discovered event GW170817. The effective one body formalism was previously shown to produce waveforms that agree with numerical simulations of aligned-spin BNS systems during the inspiral. This model includes aligned-spin effects, the quadrupolar and octopolar adiabatic tides, and the effect of dynamical tides. We improve this model by adding the spin-induced quadrupole moment and a prescription for tapering the end of the waveform to qualitatively agree with numerical simulations of the merger. The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations. We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression to efficiently sample the parameter space. The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz. With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).

@@

\red{Choose point-mass or point-particle.}

\red{Use $\chi$ for dimensionless spin instead of $S$.}

\red{Choose notation $\Lambda_2^A$ or $\Lambda_1$.}

\red{only use texttt for actual programs and functions. TaylorF2 should usually be standard text}

@@

%The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors, has demonstrated that the parameters of the binary can be measured with enough precision to answer fundamental physics questions such as identifying BNS mergers as a source of gamma ray bursts [[cite]], measuring the Hubble constant~\cite{GW170817Hubble}, and understanding the structure of neutron stars (NSs) through tidal interactions~\cite{GW170817}. Measuring these parameters accurately will also be necessary to properly interpret the gamma ray burst [[cite]] and optical afterglow [[cite]] that was found in conjunction. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} indicates that, when second generation detectors reach design sensitivity, including KAGRA~\cite{Somiya2012} and possibly LIGO-India~\cite{IyerSouradeepUnnikrishnan2011}, we could eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}. This will allow us to estimate the NS mass distribution [[cite]], spin distribution [[cite]], and universal equation of state (EOS) that is common to all NSs\cite{DelPozzoLiAgathos2013, LackeyWade2015}.

%\red{Purpose: Intro. Justify this field of research.}

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors has demonstrated that tidal interactions during the inspiral can be measured with enough precision to constrain the radius and equation of state~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, BNSPE, EOSPaper} of NSs. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be stacked to significantly improve measurements of the radius and EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}. 
%\red{[[Mention A+ being funded? How will this improve each parameter?]]}

%These results are obtained using a Bayesian analysis that compares the data to parameterized waveform models. To avoid systematic errors highly accurate models are needed, and they must include all relevant physical effects. This includes descriptions of the point-particle interactions, spins, and tides. Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models. These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5PN and 6PN order. Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Alternative models exist with more accurate representations of the point particle terms. \MP{What are these alternative models besides EOB?} However, they do not always simultaneously include the spin and tidal effects as well. Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}. In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].

%\red{Purpose: Explain the problem. Fast and accurate waveforms are needed}

These results are obtained using Bayesian analyses that compare the data to parameterized waveform models. Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. Depending on the model, this can take seconds to hours for long BNS waveforms, so analytic frequency-domain approximations are often used instead.

%In addition, to avoid systematic errors, highly accurate models are needed that include all relevant physical effects. This includes descriptions of the point-particle interactions, spins, and tides. Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models [[cite some examples]]. These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5PN and 6PN order. Unfortunately, the lack of point-particle terms above 3.5PN order can lead to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. \red{[[Describe what was found for GW170817 in PE paper? This seemed to be the case for large tidal parameters but not for the actual event. Also you didn't compare different PN waveforms there.]]}

%\red{Purpose: Explain what waveforms have been used for GW170817 and other works and why yet another one is needed.}

In addition, to avoid systematic errors that can be as large as statistical errors, highly accurate waveform models are needed that include all relevant physical effects, including point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. For GW170817, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions, on the other hand, were included simply by adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was used for the phase correction~\cite{DietrichBernuzziTichy2017}.

%For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates~\cite{PEPaper}. The first was the analytic PN waveform model known as \texttt{TaylorF2}~\cite{BuonannoIyerOchsner2009}. Because of the known systematic errors that result from PN waveforms, three additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides}, \texttt{SEOBNRv4\_ROM\_NRTides}, and \texttt{IMRPhenomP\_NRTides}. These three models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017, NewNRTidalPaper}. At low frequencies, the tidal effect is that of \texttt{TaylorF2}. It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins. The point-particle and spin prescriptions were different. For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model~\cite{HusaKhanHannam2016, KhanHusaHannam2016}. It uses \texttt{TaylorF2} at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies. The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM}~\cite{Puerrer2015} for the EOB waveform \texttt{SEOBNRv4} for BBH systems~\cite{Bohe:2016gbl}. Finally, the \texttt{IMRPhenomP\_NRTides} is an extension of \texttt{IMRPhenomD\_NRTides}, but also includes a spin-induced quadrupole moment and an approximates precession with a single effective spin parameter~\cite{}. These models are improvements to \texttt{TaylorF2}, but only treat matter effects as an additive phase correction to BBH waveform models. \red{[[Are all these details needed?]]} [[These waveform models were in broad agreement with each other for $\tilde\Lambda$ (and other parameters?) to 10\% error.]]

%We therefore build a fast surrogate that reproduces as accurately as possible \texttt{SEOBNRv4T} in the frequency domain.

%********************

%Alternative models exist with more accurate representations of the point particle terms. \MP{What are these alternative models besides EOB?} However, they do not always simultaneously include the spin and tidal effects as well. Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}. In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].

%The alternative version, \texttt{TEOBResumS}~\cite{Nagar2017}, uses a slightly different point-particle and spin prescription. And, in addition, includes the $\ell = 2, 3$, and 4 tidal effects and the spin induced quadrupole moment, but does not include the effect of dynamical tides from the induced $f$-mode resonances. It does, however, use a method for re-summing the expressions for tidal interactions.

%In addition to accuracy, useful waveform models must be fast enough to be used in Bayesian parameter estimation algorithms such as Markov chain Monte Carlo (MCMC) and Nested Sampling that currently require $10^7$--$10^8$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. Most time-domain waveform implementations such as the PN approximants and the EOB models involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. For PN waveform models, this can take several seconds for a starting frequency of 10~Hz, while for the \texttt{SEOBNRv4T} model discussed here, the equivalent evaluation time can be a day.

%For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates. The first was the stationary phase approximation PN waveform model known as \texttt{TaylorF2}~\cite{BuonannoIyerOchsner2009}. Because of the known systematic errors that result from PN waveforms, two additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides} and \texttt{SEOBNRv4\_ROM\_NRTides}. Both of these models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017}. At low frequencies, the tidal effect is that of \texttt{TaylorF2}. It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins. The point-particle and spin prescriptions were different. For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model~\cite{HusaKhanHannam2016, KhanHusaHannam2016}. It uses \texttt{TaylorF2} at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies. The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM}~\cite{Puerrer2015} for the EOB waveform \texttt{SEOBNRv4} for BBH systems~\cite{Bohe:2016gbl}. Both of these models are improvements to \texttt{TaylorF2}, but do not incorporate all known physical effects in a consistent way. We therefore build a fast surrogate that reproduces as accurately as possible \texttt{SEOBNRv4T} in the frequency domain.

%\red{[[Purpose: The EOB waveform that's expensive but solves all of humanity's problems.]]}

In this paper we will use a new implementation of the effective one body (EOB) formalism, that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note an alternative model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a slightly different point-particle and spin prescription. For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances or a tapering to zero amplitude after merger. 
%\red{[[Must mention evaluation time to segue into next 2 paragraphs.]]}

%One such alternative to these waveforms that do treat matter effects consistently with the other parameters is the effective one body (EOB) formalism discussed in section~\ref{sec:eob}. It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model. Recently two versions have become available that include both tidal effects and aligned spin and are implemented under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}. The \texttt{SEOBNRv4T} model~\cite{Hinderer:2016eia,Steinhoff:2016rfi} used in this paper, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2017}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. The other model, \texttt{TEOBResumS}~\cite{Nagar2017, NewPaper2018}, uses a slightly different point-particle and spin prescription. For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances or a tapering to zero amplitude after merger.

%\red{[[Purpose: Why other attempts for using EOB with parameter estimation have been insufficient.]]}

Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible. For GW170817~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. Unfortunately, while the results were broadly consistent with the MCMC results when using the same waveform models, there were noticeable differences. In particular the shape of the PDF was different, and the credible intervals for the tidal parameters were shifted downward by $\sim 10$\%~\cite{BNSPE}. Another limitation is that RapidPE does not currently marginalize over calibration errors. As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. In addition, different methods of formulating the problem can allow for faster integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

%*********** Part 2 of intro: Surrogates ************

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for rapidly evaluating waveform models. The essential idea is to construct a fitting function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions.
% whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. 
The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
%method that selects a subset of training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}. 
In this work, instead of using a reduced basis, we will instead use the fact that the Fourier transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we can efficiently represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

The final step in building a surrogate is to interpolate between waveform parameters $\bx$. Previous work focused on parameter spaces with three or fewer dimensions. This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality: for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the EOB waveform here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible. Recent work on optimally choosing waveforms for analytic~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

@@

This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}. 
%The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.

We organize the paper as follows. In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In Section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section IV. In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817. Finally, we discuss applications and future improvements in Section V.

@@

In the compact binary system we consider, the two material bodies, e.g. two neutron stars, with masses and spins $m_{A}$ and $\bs{S}_{A}$ for $A=1,2$, have spin-induced mass quadrupole moments $Q_{A,\mr{SS}}^{ij}=-\kappa_A S_A^{\langle i}S_A^{j\rangle}/m_A$ \cite{Poisson1998}, with $\kappa_{A}$ the quadrupole-monopole parameter. The case of two black holes is recovered for $\kappa_1=\kappa_2=1$, while for a neutron star $\kappa$ can be larger, of order 10 for hard equations of state.

>>>>

\title{Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression}

@@

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817. We present a completed version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}. This model includes the quadrupolar and octopolar adiabatic and dynamical tides, and we improve it by including the spin-induced quadrupole moment effect and a prescription for tapering the end of the waveform to agree with numerical simulations. The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. 
%This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. 
Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations. We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression. The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz. With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).

@@

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can accurately measure quantities such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be stacked to significantly improve measurements of NS radii and the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.

The reliability of these measurements, however, depend on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data. Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. Depending on the model, this can take seconds to hours for long BNS waveforms, so analytic frequency-domain approximations are often used instead.

For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

In this paper, we will use an implementation of the effective one body (EOB) formalism that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note an alternative model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a slightly different point-particle and spin prescription. For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.

Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible. In Ref.~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. Unfortunately, direct comparisons between this code and traditional samplers have not been published for BNS systems using the same waveform models. It is therefore not clear how accurate this code is at measuring quantities such as the tidal parameters. As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for rapidly evaluating waveform models. The essential idea is to construct a fitting function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions.
The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
In this work, instead of using a reduced basis, we will instead use the fact that the Fourier transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we can efficiently represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

The final step in building a surrogate is to interpolate between waveform parameters $\bx$. Many previous works focused on parameter spaces with three or fewer dimensions. This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality: for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the EOB waveform here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible. Recent work on optimally choosing waveforms for analytic~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

@@

This work also takes a different approach from the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.

%This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.

We organize the paper as follows. In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In Section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section IV. In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817. Finally, we discuss future improvements in Section VI. We also give the expressions for the TaylorF2 base model in the Appendix.

@@

In the compact binary system we consider, the two material bodies, e.g. two neutron stars, with masses and spins $m_{A}$ and $\bs{S}_{A}$ for $A=1,2$, have spin-induced mass quadrupole moments $Q_{A,\mr{SS}}^{ij}=-\kappa_A S_A^{\langle i}S_A^{j\rangle}/m_A$ \cite{Poisson1998}, with $\kappa_{A}$ the quadrupole-monopole parameter. The case of two black holes is recovered for $\kappa_1=\kappa_2=1$, while for a neutron star $\kappa$ can be larger, of order 10 for hard EOSs.
